On Optimality of Finite Element Integration

### 2015-10-18 TOMS_Paper/assembly.tex
Changed quite a lot of stuff

\subsection{Transformation Space - Introducing Pre-evaluation}
\label{sec:pre-evaluation}
For finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

@@

Pre-evaluation opportunities can be exposed through an exploration of the expression tree transformation space. This would be challenging if we were to deal with arbitrary loop nests and expressions, while we will see that a simpler algorithm can be adopted for the class arising in finite element integration. In any case, however, the following issues should be considered:
\begin{itemize}
\item In contrast to what happens with hoisting in outer product loop nests, the temporary variable size is proportional to the number and trip counts of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L N O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual execution time.
\item The transformations exposing pre-evaluation opportunities could increase the arithmetic complexity (e.g., expansion may increase the operation count; further examples will be provided later). This could outweigh the gain due to pre-evaluation.
\end{itemize}

To summarize, so far we have highlighted four problems:
\begin{enumerate}
\item The need for an algorithm to expose pre-evaluation opportunities
\item Potential explosion in working set size
\item Potential increase in operation count due to manipulation of the expression tree
\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination could inhibit pre-evaluation, while pre-evaluation might expose further sharing
\end{enumerate}

We expand on these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2), we provide a more refined definition of optimality.

@@

Sharing elimination and pre-evaluation (and, as we will clarify, their composition) allow us to explore all ways in which spatial and temporal redundancy can be eliminated from loop nests and expressions as in Definition~\ref{def:fem-loopnest}.

@@

Based on these considerations, we establish the following memory constraints:
\begin{enumerate}
\item the size of a temporary due to code motion cannot be larger than that of the outer product loop nest iteration space;
\item the total amount of memory occupied by the temporaries due to code motion cannot exceed a certain threshold, \texttt{$T_H$}.
\end{enumerate}
A corollary of constraint 1) is that hoisting expressions involving geometry-dependent terms outside of $\Lambda$, which is a possibility we discussed at the beginning of this section, is now forbidden. This clearly shrinks the transformation space, pruning points that most likely would result in sub-optimal performance.

\subsection{Reaching Optimality}
We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality}.

@@

To partly answer this question, we make use of a result -- the foundation of tensor contraction mode (Section~\ref{sec:tc}) -- from~\citeN{Kirby:TC}. Essentially, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors (see Section~\ref{sec:background}). We interpret this result at the loop nest level: with an input as in Figure~\ref{code:loopnest}, we can always dissect \texttt{F} into distinct sub-expressions (the monomials). Each sub-expression is factorizable so as to split constants from $[L_i, L_j, L_k]$-dependent terms. These $[L_i, L_j, L_k]$-dependent terms can be hoisted outside of $\Lambda$ and stored into temporaries. As part of this process, the reduction induced by $L_i$ is evaluated. Consequently, $L_i$ disappears from $\Lambda$. This transformation is actually what we were referring to when we more informally introduced pre-evaluation in Section~\ref{sec:optimal-impl}.

@@

Because of symmetries in basis functions, pre-evaluation could produce identical tables, which would be mapped to the same temporary. The transformed sub-expression (monomial) can therefore be characterized by the presence of sharing. This is why the last two lines of the algorithm, which goal is to eliminate sharing following the procedure described in Proposition~\ref{prop:optimal-mln}, are transparently applied to all sub-expressions, regardless of whether pre-evaluation was used. The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the usual input of Figure~\ref{code:loopnest}.

>>>>

\subsection{Transformation Space - Pre-evaluation}
\label{sec:pre-evaluation}
For finite element integration loop nests, the presence of sharing is a \textit{sufficient but not necessary} condition for being in \textit{non-optimal} form. Consider again the bilinear form implementation in Figure~\ref{code:loopnest}. We pose the following question: are we able to identify sub-expressions within $F$ for which the reduction imposed by $L_i$ can be pre-evaluated, thus obtaining a decrease in operation count proportional to the size of $L_i$, $M$? The transformation we look for is exemplified in Figures~\ref{code:loopnest_red} and~\ref{code:loopnest_nored}; Figure~\ref{code:loopnest_red}, which represents the input, can be seen as a simple instance of the abstract loop nest in Figure~\ref{code:loopnest}.

@@

Pre-evaluation opportunities can be exposed through an exploration of the expression tree transformation space. This would be challenging if we were to deal with arbitrary loop nests and expressions. However, we can make use of a result -- the foundation of tensor contraction mode -- to build a relatively simple algorithm. As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces; such monomials can always be reduced to a product of two tensors. This result can be turned into a loop and expression transformation algorithm, for which we provide a succinct description below.

\begin{Algo}[Pre-evaluation]
\label{algo:pre-evaluation}
\normalfont
Consider a finite element integration loop nest $\Lambda = [L_e, L_i, L_j, L_k]$ as in Figure~\ref{code:loopnest}. We dissect \texttt{F} into distinct sub-expressions (the monomials). Each sub-expression is factorized so as to split constants from $[L_i, L_j, L_k]$-dependent terms. This transformation is feasible, as a consequence of the results in~\citeN{Kirby:TC}. These $[L_i, L_j, L_k]$-dependent terms are hoisted outside of $\Lambda$ and stored into temporaries. As part of this process, the reduction induced by $L_i$ is evaluated. Consequently, $L_i$ disappears from $\Lambda$. 
\end{Algo}

The pre-evaluation of a monomial introduces, however, some critical issues:
\begin{itemize}
\item In contrast to what happens with hoisting in outer product loop nests, the temporary variable size is proportional to the number and trip counts of non-reduction loops crossed (for the bilinear form implementation in Figure~\ref{code:loopnest}, $N O$ for sub-expressions depending on $[L_i, L_j, L_k]$ and $L N O$ for those depending on $[L_e, L_i, L_j, L_k] $). This might shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual execution time.
\item The transformations exposing $[L_i, L_j, L_k]$-dependent terms increase, in general, the arithmetic complexity (e.g., expansion may increase the operation count). This could outweigh the gain due to pre-evaluation.
\item The need for a strategy to coordinate sharing elimination and pre-evaluation opportunities: sharing elimination could inhibit pre-evaluation, while pre-evaluation might expose further sharing
\end{itemize}

We address these points in Section~\ref{sec:optimal-synthesis}. In the perspective of addressing point 2), we conclude providing a more refined definition of optimality.

@@

Sharing elimination and pre-evaluation allow us to construct an algorithm to explore all ways in which spatial and temporal redundancy can be eliminated from loop nests and expressions as in Definition~\ref{def:fem-loopnest}.

@@

Based on these considerations, we establish the following memory constraints.

\begin{Const}
\label{const:Le}
The size of a temporary due to code motion must not be proportional to the size of $L_e$.
\end{Const}

\begin{Const}
\label{const:TH}
The total amount of memory occupied by the temporaries due to code motion must not exceed a certain threshold, \texttt{$T_H$}.
\end{Const}

Constraint~\ref{const:Le} reflects the policy decision that the compiler should not silently consume memory on ... data objects. Consequently, generalized code motion as performed by sharing elimination is not allowed to hoist expressions outside of $L_e$.

\subsection{Reaching Optimality}
We here discuss how we can systematically reach the goal set by Definition~\ref{def:optimality} under Constraints~\ref{const:Le} and~\ref{const:TH} for the class of loop nests in Definition~\ref{def:fem-loopnest}.

@@

@@

Because of reuse of basis functions, pre-evaluation may produce identical tables, which will be mapped to the same temporary. The transformed sub-expression (monomial) can therefore be characterized by the presence of sharing.

The output of the transformation algorithm is as in Figure~\ref{code:loopnest-opt}, assuming the usual input of Figure~\ref{code:loopnest}.
